---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
output: fost::diabetologia_manuscript
---

# Title page

Title: Association of non-esterified fatty acid composition with insulin
sensitivity and beta cell dysfunction in the Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

<!-- 47 characters + spaces -->
<!--Running title: NEFA and the pathogenesis of diabetes-->

<!-- First, middle initial, last -->
Author: Luke W. Johnston^1^; Stewart B. Harris^2^; Ravi Retnakaran
^3,4^; Adria Giacca^5^; Zhen Liu^1^;
Richard P. Bazinet^1^; and Anthony J. Hanley^1,3,6^

Institutions of origin:

 (1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
 (2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
 (3) Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
 (4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
 (5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
 (6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.

Corresponding author:

- Name: Anthony J. Hanley
- Current address:  
    Department of Nutritional Sciences  
    Faculty of Medicine  
    University of Toronto  
    FitzGerald Building, 150 College Street, Room 341  
    Toronto, ON, Canada, M5S 3E2
- Phone number: 416.978.3616
- Email: anthony.hanley@utoronto.ca

Word count - Main text: 3869 / 4000 (with figure captions)

<!-- Table and figure count: 4 / 4 -->

<!-- Reference limit 49 / 50 -->

# Abstract

Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on the pathogenesis of diabetes, specifically on differences in the insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.

Methods: In the PROMISE longitudinal cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years. Outcome variables were calculated using the OGTT values. Insulin sensitivity was assessed using HOMA2-%S and the Matsuda index, while beta cell function was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were used, adjusting for time, waist, sex, ethnicity, baseline age, alanine aminotransferase, and physical activity. NEFA were analysed as both concentrations (nmol/mL) and proportions (mol%) of the total fraction.

Results: Participants (73% female, 70% with European-ancestry) had their insulin sensitivity and beta cell function decline by 14-27% over 6 years of follow-up. While in unadjusted models there were several NEFA (e.g. 18:1n-7, 20:1n-9, 22:4n-6) that had negative associations with insulin sensitivity, nearly all of these associations were attenuated in fully adjusted models. In adjusted models, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive associations with HOMA2-%S.

Conclusions: Total NEFA concentration was a strong predictor of lower beta cell function over 6 years. Our results suggest that the association with beta cell function is due to the absolute size of the serum NEFA fraction, rather than the specific fatty acid composition.

## Keywords:

beta cell function; diabetes pathogenesis; fatty acid composition; insulin sensitivity; longitudinal cohort; non-esterified fatty acids; 

## Abbreviations:

- AR1: Autoregressive of order 1
- ALT: Alanine aminotransferase
- EPA: Eicosapentaenoic acid
- FDR: False Discovery Rate
- GEE: Generalized estimating equations
- HOMA: Homeostatic model assessment
- HOMA-IR: Homeostatic model assessment of insulin resistance
- HOMA2-%S: Homeostatic model assessment 2 of insulin sensitivity
- IFG: Impaired fasting glucose
- IGI/IR: Insulinogenic index over HOMA-IR
- IGT: Impaired glucose tolerance
- ISI: Insulin Sensitivity Index (Matsuda index)
- ISSI-2: Insulin Secretion-Sensitivity Index-2
- MAQ: Modified Activity Questionnaire
- MET: Metabolic Equivalent of Task
- NEFA: Non-esterified fatty acids
- OGTT: Oral glucose tolerance test
- PLS: Partial least squares regression
- PROMISE: Prospective Metabolism and Islet Cell Evaluation Cohort
- PUFA: Polyunsaturated fatty acids
- QIC: Quasi-Likelihood Information Criterion
- TAG: Triacylglyceride
- WC: Waist circumference

```{r setup, include=FALSE}
devtools::load_all('.')
set_options()
```

```{r captions}
## Include captions below using `captioner` package
library(captioner)
fig <- captioner(prefix = 'Fig.')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner(prefix = 'ESM Fig.')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner(prefix = 'ESM Table')
cite_st <- pryr::partial(stab, display = 'cite')

tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over 6 years in the PROMISE cohort. Models are only adjusted for time. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
```

# Introduction

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and can impair 
pancreatic beta cell production of insulin [@Rachek2014a;@Cnop2008a]. 
Observational and clinical studies have reported concordant findings, showing in
particular that elevated total plasma NEFA associates with an increased risk for
incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes 
utilised individual fatty acids such as palmitic acid (16:0) or oleic acid 
(18:1n-9), or alternatively used specific oils such as soybean oil, which is 
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the 
exposure to characterise the metabolic impact of total NEFA. However, fatty
acids comprise multiple molecules with diverse physiological functions, and few 
studies have analysed the effects of a broader spectrum of fatty acids. For instance,
one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta cells [@Kato2008].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], have
been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have been
epidemiological studies that only looked at total NEFA and not individual fatty 
acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no
longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted
natural history of diabetes and the growing appreciation of the divergent
effects of individual fatty acids. Therefore, our objective was to examine the
association of serum NEFA composition on differences over time in insulin
sensitivity and beta cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as eicosapentaenoic acid in the
NEFA fraction would associate with declining insulin sensitivity and beta cell
function over 6 years.

# Methods

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
Participants are contacted annually by telephone. The
current study sample used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterisation, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardised across the 2 centres and research
nurses were centrally trained. 

## Blood measure assessments

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyser and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorised using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualised under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given their
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see ref @Johnston2016a for the
analysis of the phospholipid and cholesteryl ester fractions).

## Outcome variables

Insulin sensitivity and beta cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA2-%S largely
reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
sensitivity [@AbdulGhani2007].  beta cell function was assessed using the
Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of
the first phase of insulin secretion while ISSI-2 is analogous to the
disposition index (but using OGTT values). Each index has been validated against
gold standard measures [@Hermans1999a;@Matsuda1999;@Retnakaran2009]. Formulas
used to calculate these measures can be found in the ESM Methods.

## Anthropometrics and sociodemographics

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Sociodemographic information, including age, sex, and ethnicity, were determined
using questionnaires administered at each examination. In the lifestyle 
questionnaire, physical activity was determined using a version of the 
Modifiable Activity Questionnaire (MAQ) [@Kriska1990]. The MAQ collects 
information on leisure and occupational activity, including intensity, 
frequency, and duration, over the past year. Each reported activity from the 
MAQ was weighted by its metabolic intensity allowing for the estimation of 
MET-hours per week.

## Statistical analysis

The primary outcome variables for this analysis were 
HOMA2-%S, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given it's capacity to
handle any missed visits and for the correlation of variables between time points. 
See the ESM Methods for more detail on GEE.
The predictor variables and continuous covariates were
scaled (mean centered and standardised). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC) [@Pan2001a]. See
the ESM Methods for more detail on QIC and the directed acyclic graphs. 
`r cite_st('qic')` shows the covariates compared using QIC. We selected the
model (M9) that had the fewest covariates and that had similar fit between the
outcome measures. The final GEE model included the covariates years from baseline
(time), sex (female or male), ethnicity (European ancestry or non-European 
ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
and baseline age were time-*independent* variables. See the ESM Methods for more
detail on the covariate inclusion. After scaling,
and log-transforming the model variables, then exponentiating the GEE estimates,
the interpretation of the GEE results is as an expected percent difference in 
the outcome variable for every standard deviation (SD) increase in the predictor
variable given the covariates are held constant. With the main effect 
associations (not with a time interaction), this interpretation is applied 
consistently at each time point and thus represents associations of baseline 
NEFA for each future value of the outcomes. Since TAG is a risk factor for
diabetes and since NEFA contribute to TAG production [@Barrows2006a], TAG may
act as a mediator between NEFA and the outcomes. To determine the role of TAG in
the association between NEFA and the outcomes, TAG was included in the GEE model
in a sensitivity analysis. Lastly for the GEE models, we tested for an 
interaction with sex, ethnicity, or time by the predictor term for each outcome 
variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyse the inherent multivariate nature of the
composition of the NEFA fraction, as each model only contains data from a single
fatty acid. Therefore, to confirm the results of the GEE analyses in a
multivariate environment (all NEFA variables in a single model), partial least
squares (PLS) was used with the outcomes as the dependent variable.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a]
and the package pls `r packageVersion('pls')` for the PLS analysis.
The R code and output for all analyses of this manuscript is available at 
https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate (FDR) [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].
The cohort datasets analysed during the current study are available from the
corresponding author on reasonable request.

# Results

```{r inline_results}
nefa <- calc_pct_nefa(project_data)
chg_outcomes <- calc_outcomes_chg_n(project_data)

ds0yr <- dplyr::filter(project_data, VN == 0)
female <- calc_npercent(ds0yr$Sex, 'Female')
european <- calc_npercent(ds0yr$Ethnicity, 'European')
famhist <- calc_npercent(ds0yr$FamHistDiab, 'Yes')
est_gee <- gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    calc_gee_magnitude()

meds_use <- table(project_data$MedsLipidsChol, project_data$VN)[2, c(1, 3)]

sig_fa_unadj <- gee_results %>% 
    dplyr::filter(model == "Unadjusted", p.value < 0.05) %>% 
    dplyr::select(Yterms, Xterms, unit) %>% 
    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
sig_fa_adj <- gee_results %>% 
    dplyr::filter(model == "Adjusted", p.value < 0.05) %>% 
    dplyr::select(Yterms, Xterms, unit) %>% 
    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
```

## Basic characteristics of the PROMISE cohort

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from further GEE analyses. There were `r meds_use[1]` and
`r meds_use[2]` participants who used lipid and cholesterol modifying medications
at the baseline and 6-year visit, respectively. Medication use did not influence
NEFA composition, the outcomes, or the GEE models (data not shown).

`r cite_f('nefa')` shows the compositional distribution of NEFA in the study
participants. Four individual NEFA made up the vast majority 
(`r sum(nefa$pct)`%) of the total NEFA fraction. The largest contributors
were `r nefa$c[1]`, `r nefa$c[2]`, `r nefa$c[3]`, and `r nefa$c[4]`. Raw
concentration values are shown in `r cite_st('nefa')`. All individual NEFA as
well as the total fraction had correlations that ranged from weak to null
(r<0.3) with participant characteristics (see `r cite_sf('corr_conc')` and 
`r cite_sf("corr_pct")`).

## GEE modeling

A number of associations were seen in the unadjusted GEE models 
(see `r cite_f('gee_unadj')`), particularly for NEFA modeled as a concentration.
There were several fatty acids (`r paste(unique(sig_fa_unadj$Xterms),collapse=", ")`)
that had significant associations with both insulin sensitivity and beta cell
function outcomes. Most associations were negative, however, only 20:5n-3
(eicosapentaenoic acid; EPA) had positive associations with insulin sensitivity
as both a mol% and as a concentration. One fatty acid, 18:1n-7 (vaccenic acid),
had consistent negative associations with all four outcomes.

After full model adjustment (`r cite_f('gee_adj')`), most significant 
associations were attenuated, particularly with insulin sensitivity. The
remaining fatty acids (`r paste(unique(sig_fa_adj$Xterms),collapse=", ")`) had
significant associations, of which only EPA had a significant association (with
HOMA2-%S) as a mol% and was the only fatty acid with a positive association. All
other fatty acids had negative associations with IGI/IR and ISSI-2. The
magnitude of association for each of these variables was fairly consistent for
each beta cell function measure. For every one SD increase in any of the NEFA
variables associated with beta cell function, there was an average predicted
`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at any
given clinic visit.

Sensitivity analyses revealed that adjustment for WC was responsible for 
attenuating the unadjusted results (see `r cite_sf("gee_adj_nowaist")`). Adjusting for TAG 
attenuated all associations with IGI/IR and ISSI-2 (data not shown), though this
may be due to the fact that only a subset of participants at the 6 year visit 
had TAG measured. Sensitivity analyses to examine the association with the sum 
of fatty acid classes (saturated vs unsaturated) revealed similar findings as 
seen with the adjusted GEE model for total NEFA (data not shown). There were no 
significant interaction effects of time, sex, or ethnicity in either unadjusted
or adjusted models. Medication use for controlling lipid and cholesterol did not
influence the results (data not shown). Raw values from the GEE models are shown
in `r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee_adj')` for
fully adjusted models. Consistent with the GEE modeling, results from modeling
all NEFA in a multivariate environment using PLS showed no discernible
clustering nor was it predictive of the outcomes (data not shown).

# Conclusions

In a Canadian population of adults who were at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta cell
function over 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for these fatty acids as a
proportion, an observation that was further confirmed by multivariate cluster
techniques, which found no ability of these individual NEFA to predict beta cell
function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
(20:5n-3), had a positive association with insulin sensitivity. In general,
these observations suggest that the absolute size of the total NEFA fraction,
rather than its specific composition (aside from EPA), likely influences the 
pathogenesis of diabetes, at least within a population at-risk for diabetes.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures over time. However, we
believe this is a strength for our specific objective, as the potential impact
that reverse causality, due to the tight integration between fatty acid and
glucose metabolism, may have on the results is lower. This is also an
observational cohort, and there may be some residual confounding we have not
considered or that could not be measured. Nonetheless, potential covariates were
empirically analysed prior to inclusion into the GEE models to best understand
and minimise potential confounding. While there may be potential overadjustment
due to controlling for WC, both QIC and directed acyclic graph outputs
recommended including WC in the GEE model. Even still, we ran sensitivity
analyses and identified WC as the covariate that attenuated the unadjusted
results. This attenuation may be due to the fact that higher NEFA may be 
reflective of greater adipose tissue mass and acting as a mediator between the 
adipose tissue and insulin sensitivity (since adipocytes are active endocrine 
organs).

Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and, while sex and ethnicity did
not influence the results presented from this cohort, our findings may not
entirely be generalisable to other populations. However, given these
limitations, our study also has several strengths, including the longitudinal
design and the rigorous statistical techniques and methods applied, which are
specifically suited to investigating temporal relationships and to handling the
multivariate nature of the data. Lastly, our cohort contains highly detailed and
comprehensive variable measurements at each collection visit, and has both
concentration and mol% data for the fatty acids.

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta cell function
[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta cell function. No fatty acid as a mol% had associations with
beta cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta cells.

Biologically, in free-living populations chronic elevation of NEFA may be 
mediating its association with metabolic outcomes through TAG. In normal 
metabolism, NEFA enters the liver and assists in the production of TAG that is
to be processed into very-low density lipoproteins (VLDL) [@Nielsen2012a]. As
such, higher NEFA may contribute to hypertriglyceridemia, which is a known risk
factor for diabetes. In the sensitivity analysis adjusting for TAG, all
associations with beta cell function were attenuated, suggesting that NEFA may
in fact be mediating its association with beta cell dysfunction through higher
TAG.

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signalling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw in general no association of any individual or total NEFA with
hepatic (HOMA2-%S) or whole-body insulin sensitivity (ISI), except for EPA. 
There is some evidence [@Steffen2015a;@Virtanen2014;@Sarbolouki2013] that EPA
may be involved with lowering the risk for diabetes, potentially due to its role
in inflammation [@Calder2009]. However, given we found the association only with
HOMA2-%S and not with ISI, the significant association could be a chance finding,
which will require future studies to confirm. We also identified WC as having a
powerful attenuating effect on the associations between unadjusted and adjusted
models. There are some possible explanations for our null findings for the
insulin sensitivity measures and the influence of WC on the results. First, WC
may be a strong causal link in previously reported associations between NEFA and
insulin sensitivity, given the role adipocytes play in metabolism (e.g. leptin,
adiponectin). Second, there may be differences in physiology between fasting and
postprandial NEFA kinetics that we are not able to investigate but that may
explain our null findings for fasting NEFA and insulin sensitivity. For
instance, some experimental studies using clamp protocols found that fasting
NEFA was a weak predictor of insulin sensitivity compared to postprandial
concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
which may be more metabolically active given postprandial activity. Third, the
null findings seen may be due to the high risk population examined in PROMISE, 
as the majority of subjects had a high WC and BMI. It may be that in this
population, insulin resistance has become well established and NEFA may not
contribute to insulin resistance at this somewhat more advanced stage in the
pathogenesis of diabetes.

Few studies have examined the composition of NEFA on metabolic functioning. One 
recent, well-analysed study used a variety of advanced fatty acid measurement 
and statistical techniques to explore the multivariate relationship between the 
NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a]. 
Specifically, the authors identified NEFA 16:1n-9, 20:1n-9, and 22:4n-6 to
correlate with components of the MetS. Another similar study examining diabetes
found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers for identifying
healthy compared to individuals with diabetes [@Liu2010a]. However, both studies were
limited by smaller sample sizes (approximately 100 subjects) and the
cross-sectional design.

In conclusion, we found that total NEFA was a strong predictor for lower 
beta cell function over 6 years, irrespective of the specific composition of the
NEFA fraction, suggesting that efforts at lowering circulating total NEFA 
through medication and/or lifestyle strategies to reduce the risk of diabetes 
are well warranted. We also found that higher EPA was associated with higher
hepatic insulin sensitivity. While future studies are needed to confirm these
findings, our results reinforce the importance of continuing to investigate the
role of circulating NEFA concentration on the natural history of diabetes.

# Acknowledgements

The authors thank J Neuman, PV Nostrand, S Kink, and A Barnie (all of the
Leadership Sinai Centre for Diabetes, Mount Sinai Hospital, Canada) and
S Porter and M Marin (both of the Centre for Studies in Family Medicine,
University of Western Ontario, Canada) for their expert technical
assistance and dedication in their work for PROMISE.

## Funding

This study was supported by grants from the Canadian Diabetes Association (CDA),
the Canadian Institutes for Health Research, and the University of Toronto
Banting and Best Diabetes Centre; LWJ was supported by a CDA Doctoral Student
Research Award; RR was supported by a Heart and Stroke Foundation of Ontario
Mid-Career Investigator Award; SBH holds the CDA Chair in National Diabetes
Management and the Ian McWhinney Chair of Family Medicine Studies at the
University of Western Ontario; RPB holds a Tier II Canada Research Chair in
Brain Lipid Metabolism; AJH holds a Tier II Canada Research Chair in Diabetes
Epidemiology.

## Duality of interest

The authors declare that there is no duality of interest associated with this
manuscript.

## Contribution statement

The authors had the following responsibility: LWJ conducted research, analysed
data, and wrote the paper; RR, ZL, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content and are the guarantors of this work. All authors read and
approved the final manuscript. 

# Tables and figures

```{r table1_basic}
table_basic(project_data, caption = tab_basic)
```

Note: Values are in median (IQR), mean (SD), and n (%). The proportion of ethnic
and sex groups did not change over the 6 years. Outcome variables sample size
ranges from `r chg_outcomes$n` over the 3 visits.

```{r figure1_nefa, fig.cap=fig_dist}
plot_nefa_distribution()
```

```{r figure2_gee_unadj, fig.cap=fig_gee_unadj}
gee_results %>% 
    dplyr::filter(model == "Unadjusted") %>% 
    dplyr::mutate(unit = ifelse(unit == "Total", " ", unit)) %>% 
    plot_gee_main()
```

```{r figure3_gee_adj, fig.cap=fig_gee_adj}
gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    dplyr::mutate(unit = ifelse(unit == "Total", " ", unit)) %>% 
    plot_gee_main()
```

# References
